Proof equivalence in MLL is PSPACE-complete

### 2014-12-10 main/LMCS.tex
"notation" subsection; separated macro file

\usepackage{mllequiv,willemtools}

@@

% MLL
%   remove naming - work with occurrences instead ??
%   explain cut-elimination
%   define *-autonomous categories

% PROOF NETS
%   separate section for proof nets
%   discuss both small-step and big-step equivalence
%      connect jumps anywhere / to positive leaves (1,a) only
%   example translation: proof => net
% 
%   add 3-partition example
%

@@

@@

\begin{remark}
A main distinction with the proof nets of \cite{Blute-Cockett-Seely-Trimble-1994} is that the latter are defined as graphs whose vertices represent inferences, and whose edges represent formulae. This formulation allows an easy inclusion of cuts, which are not present in our proof nets. That the jumps from $\bot$-vertices in these graphs are attached to edges, not vertices, may seem curious at first, but it is a natural choice given the categorical view of $\bot$-introduction as the isomorphism $A\to A\pr\bot$.
\end{remark}

@@

A main technical advantage of this restriction is that cut-elimination can be performed in a one-shot operation, by path composition over jumps. Without the need to define cut-elimination by local transformations, cuts may then be omitted from proof nets altogether, which justifies the present definition.

@@

\begin{proposition}[\cite{Blute-Cockett-Seely-Trimple-1994,Hughes-2012-nets}]
\label{prop:proof nets work}
%
If $\Pi\toNet\links$ and $\Pi'\toNet\links'$, then $\Pi\perm\Pi'$ if and only if $\links\perm\links'$.
%
\end{proposition}

@@

\begin{figure}
\[
	\netA 1223 ~\perm*~
	\netA 1323 ~\perm*~
	\netA 1321 ~\perm*~
	\netA 2321
\]
\caption{A rewiring sequence on proof nets}
\label{fig:net rewiring}
\end{figure}

\subsection*{Notation}

We will use a concise diagrammatic notation for sequents and proof nets.
%
The units $1$ and $\bot$ are represented by a circle~($\circ$) and a disc~($\bullet$) respectively;
%
formulae related by a tensor will be connected by edges; 
formulae related by a par will be juxtaposed, and collected in a box when a par-formula is an immediate subformula of a tensor-formula (and also for illustrative purposes).
%
For example, the following denote the same sequent:
\[
	\bot\tn\bot,\bot\tn\bot,\1,(\1\pr\1\pr\1)\tn\bot
\]
\[
\begin{tikzpicture}[x=5mm,y=-5mm,octo]
	\draw (0,0) node[bullet] (a) {} -- (1,0) node[bullet] (b) {} (2,0) node[bullet] (c) {} -- (3,0) node[bullet] (d) {};
	\node[circ] (1) at (4,0) {};
	\node[circ] (2) at (5.25,0) {}; \node[circ] (3) at (6.25,0) {}; \node[circ] (4) at (7.25,0) {};
	\draw[rounded corners] (4.75,-1) rectangle (7.75,1);
	\draw (7.75,0) -- (8.75,0) node[bullet] (e) {};
\end{tikzpicture}
\]
The links of a proof net are added to the sequent as coloured arrows.
%
The following example is a proof net for the above sequent.
\[
\begin{tikzpicture}[x=5mm,y=-5mm,octo]
	\node[circ] (1) at (6,0) {};
	\draw (0,0) node[bullet] (a) {} -- (1,0) node[bullet] (b) {} (3,0) node[bullet] (c) {} -- (4,0) node[bullet] (d) {};
	\node[circ] (2) at (1,3) {}; \node[circ] (3) at (2,3) {}; \node[circ] (4) at (3,3) {};
	\draw[rounded corners] (0,2) rectangle (4,4);
	\draw (4,3) -- (5,3) node[bullet] (e) {};
	\begin{scope}[thepink,->]
			\draw[bend right=10] (a) to (2.center);
			\draw[bend right=10] (b) to (3.center);
			\draw[bend left=10]  (c) to (3.center);
			\draw[bend left=10]  (d) to (4.center);
			\draw[bend right=10] (e) to (1.center);
	\end{scope}
\end{tikzpicture}
\]

>>>>

\usepackage{LMCS_macros,willemtools}

@@

% MLL
%   x remove naming - work with occurrences instead ??
%   v explain cut-elimination
%     define *-autonomous categories

% PROOF NETS
%   v separate section for proof nets
%   v discuss both small-step and big-step equivalence
%      connect jumps anywhere / to positive leaves (1,a) only
%     example translation: proof => net
% 
%     add 3-partition example
%

@@

An example is given in Figure~\ref{fig:example trans}, where a proof is interpreted as the proof net of Figure~\ref{fig:example net}, by indicating the chosen jumps in the conclusion of each $\bot$-inference. (In the example, to trace formulae through inferences, the five $\bot$-formulae are kept in order from left to right, while the three $1$-formulae have explicit names.)

\begin{figure}
\[
	\exampletrans
\]
\caption{Interpreting a proof as the proof net of Figure~\ref{fig:example net}}
\label{fig:example trans}
\end{figure}

@@

\begin{remark}
A main distinction with the proof nets of \cite{Blute-Cockett-Seely-Trimble-1994} is that the latter are defined as graphs whose vertices represent inferences, and whose edges represent formulae. This formulation allows an easy inclusion of cuts, which are not present in our proof nets. That the jumps from $\bot$-vertices in these graphs are attached to edges, not vertices, is a natural choice given the categorical view of $\bot$-introduction as the isomorphism $A\to A\pr\bot$.
\end{remark}

@@

A main technical advantage of this restriction is that cut-elimination can be performed in a one-shot operation, by path composition over jumps. Then, without the need to define cut-elimination by local transformations, cuts may be omitted from proof nets altogether, which justifies the present definition.

@@

\begin{proposition}[\cite{Blute-Cockett-Seely-Trimble-1994,Hughes-2012-nets}]
\label{prop:proof nets work}
%
If $\Pi\toNet\links$ and $\Pi'\toNet\links'$, then $\Pi\perm\Pi'$ if and only if $\links\perm\links'$.
%
\end{proposition}

\subsection*{Diagram notation}

To concisely represent larger sequents and proof nets, we will introduce a compact diagrammatic notation.
Briefly, the units $1$ and $\bot$ are represented by a circle~($\circ$) and a disc~($\bullet$) respectively, a tensor $(\tn)$ will be indicated by an edge, and a par $(\pr)$ will be indicated by a box.
For example, the sequent below left is represented as a diagram below right.
\[
	\bot\tn\bot~,~\bot\tn\bot~,~\1~,~(\1\pr\1\pr\1)\tn\bot
\qquad\qquad
\vc{\begin{tikzpicture}[x=5mm,y=-5mm,octo]
	\draw (0,0) node[bullet] (a) {} -- (1,0) node[bullet] (b) {} (2,0) node[bullet] (c) {} -- (3,0) node[bullet] (d) {};
	\node[circ] (1) at (4,0) {};
	\node[circ] (2) at (5.25,0) {}; \node[circ] (3) at (6.25,0) {}; \node[circ] (4) at (7.25,0) {};
	\draw[rounded corners] (4.75,-1) rectangle (7.75,1);
	\draw (7.75,0) -- (8.75,0) node[bullet] (e) {};
\end{tikzpicture}}
\]
The salient features of a diagram are the connectedness by $\tn$-edges and containment within a $\pr$-box; respecting these, diagrams may be laid out in the plane in any way that is convenient. More formally, then, sequents and formulae are interpreted as diagrams as follows.
\[
\begin{array}{c@{~}c@{~}c}
	\bot & \Rightarrow & \bullet \\[5pt]
	1	 & \Rightarrow & \circ
\end{array}
\qquad
\begin{array}{rcll}
	A_1\tn A_2\tn\dotso\tn A_n 
	& \Rightarrow &
	  \vc{\begin{tikzpicture}[octo]
	    \node[draw,rounded corners] (1) at (0,0) {$A_1$};
	    \node[draw,rounded corners] (2) at (1,0) {$A_2$};
	    \node[draw,rounded corners] (n) at (3,0) {$A_n$};
	    \node at (2,0) {$\dots$};
	    \draw (1)--(2)--(1.6,0) (2.4,0)--(n);
	    \path[use as bounding box] (-.5,-.5) rectangle (3.5,.5);
	  \end{tikzpicture}}
	& (A_i~\text{not a $\tn$-formula})
\\[5pt]
	A_1\pr A_2\pr\dotso\pr A_n
	& \Rightarrow &
	  \vc{\begin{tikzpicture}[octo]
	    \node[draw,rounded corners] at (0,0) {$A_1$};
	    \node[draw,rounded corners] at (1,0) {$A_2$};
	    \node[draw,rounded corners] at (3,0) {$A_n$};
	    \node at (2,0) {$\dots$};
	    \draw[rounded corners] (-.5,-.5) rectangle (3.5,.5);
	  \end{tikzpicture}}
	& (A_i~\text{not a $\pr$-formula})
\\[5pt]
	A_1, A_2,\dotsc, A_n
	& \Rightarrow &
	  \vc{\begin{tikzpicture}[octo]
	    \node[draw,rounded corners] at (0,0) {$A_1$};
	    \node[draw,rounded corners] at (1,0) {$A_2$};
	    \node[draw,rounded corners] at (3,0) {$A_n$};
	    \node at (2,0) {$\dots$};
	    \path[use as bounding box] (-.5,-.5) rectangle (3.5,.5);
	  \end{tikzpicture}}
	& \quad \text{where }A_i~\Rightarrow~
	  \vc{\begin{tikzpicture}[octo]
	    \node[draw,rounded corners] {$A_i$};
	  \end{tikzpicture}}
\end{array}
\]
In the diagrammatic notation, as in the regular notation before, the jumps of a proof net are added to the sequent as coloured arrows. Per the restriction mentioned previously, jumps always connect a $\bot$-formula to a $1$-formula. The example below gives a proof net for the above sequent, presented in both notational styles.
\[
\vc{\begin{tikzpicture}[x=6mm,y=6mm,inner sep=1pt]
	\node (a) at (0,1) {$\bot$};
	\node (b) at (1,0) {$\tn$};
	\node (c) at (2,1) {$\bot$};
	\node (d) at (3,1) {$\bot$};
	\node (e) at (4,0) {$\tn$};
	\node (f) at (5,1) {$\bot$};
	\node (g) at (6,0) {$1$};
	\node (h) at (7,3) {$1$};
	\node (i) at (8,2) {$\pr$};
	\node (j) at (9,1) {$\pr$};
	\node (k) at (9,3) {$1$};
	\node (l) at (10,0) {$\tn$};
	\node (m) at (10,2) {$1$};
	\node (n) at (11,1) {$\bot$};
	\node at (0,-.5) {};
	\draw[thick] (a)--(b)--(c) (d)--(e)--(f) (j)--(l)--(n);
	\draw[thick,dashed] (h)--(i)--(k) (i)--(j)--(m);
	\draw[jump] (a) -- +(0,2) to[out=90,in=90] (h);
	\draw[jump] (c) -- +(0,2) to[out=90,in=90] (k);
	\draw[jump] (d) -- +(0,1.5) to[out=90,in=120] (k);
	\draw[jump] (f) -- +(0,1.5) to[out=90,in=90] ($(m)+(0,0.5)$) -- (m);
	\draw[jump] (n) -- +(0,2.5) to[out=90,in=90] ($(g)+(0,3.5)$) -- (g);
\end{tikzpicture}}
\qquad\qquad
\vc{\begin{tikzpicture}[x=5mm,y=-5mm,octo]
	\node[circ] (1) at (6,0) {};
	\draw (0,0) node[bullet] (a) {} -- (1,0) node[bullet] (b) {} (3,0) node[bullet] (c) {} -- (4,0) node[bullet] (d) {};
	\node[circ] (2) at (1,3) {}; \node[circ] (3) at (2,3) {}; \node[circ] (4) at (3,3) {};
	\draw[rounded corners] (0,2) rectangle (4,4);
	\draw (4,3) -- (5,3) node[bullet] (e) {};
	\begin{scope}[thepink,->]
			\draw[bend right=10] (a) to (2.center);
			\draw[bend right=10] (b) to (3.center);
			\draw[bend left=10]  (c) to (3.center);
			\draw[bend left=10]  (d) to (4.center);
			\draw[bend right=10] (e) to (1.center);
	\end{scope}
\end{tikzpicture}}
\]
The above examples illustrate how the diagrammatic notation replaces the connectives by geometric features, to allow for more convenient placement in the plane. The diagrams preserve the connectedness of non-switched edges, while switched components are placed within a box, so that a switching graph is obtained by replacing each box by an edge to exactly one of its components.

To lay out a sequent as a diagram in two dimensions may lead to ambiguity when multiple occurrences of the same formula are present, as is the case for the formula $\bot\tn\bot$ in the above examples. Where necessary, this will be resolved by adding the names of $\bot$-formulae and $1$-formulae to a diagram. When discussing different proof nets over the same sequent, the sequent will be laid out in a consistent manner.

Figure~\ref{fig:_*_,1,1,1,_*_} uses the diagrammatic notation to display the rewiring relation over the proof nets for the sequent $\bot\tn\bot,1,1,1,\bot\tn\bot$, an interesting example also highlighted in \cite{Lamarche-Strassburger-2005}. There are 24 proof nets in total, forming two equivalence classes, each consisting of a single cyclic rewiring path.

\begin{figure}
\[
\begin{array}{@{}ccccccccccc@{}}
	\netA 1223 &\perm*& \netA 1323 &\perm*& \netA 1321 &\perm*& \netA 2321 &\perm*& \netA 2331 &\perm*& \netA 2131 \\ \\ \perm* &&&&&&&&&& \perm* \\ \\
	\netA 1213 &\perm*& \netA 3213 &\perm*& \netA 3212 &\perm*& \netA 3112 &\perm*& \netA 3132 &\perm*& \netA 2132 
	\\ \\ \\ \\ \\
	\netA 1232 &\perm*& \netA 1332 &\perm*& \netA 1312 &\perm*& \netA 2312 &\perm*& \netA 2313 &\perm*& \netA 2113 \\ \\ \perm* &&&&&&&&&& \perm* \\ \\
	\netA 1231 &\perm*& \netA 3231 &\perm*& \netA 3221 &\perm*& \netA 3121 &\perm*& \netA 3123 &\perm*& \netA 2123
	\\ \\
\end{array}
\]
\caption{All proof nets for the sequent $\bot\tn\bot,1,1,1,\bot\tn\bot$}
\label{fig:_*_,1,1,1,_*_}
\end{figure}

\section{Proof search complexity}

% TO DO:
%	n-fold bottoms & ones (below)
%   3-sat reduction for proof search

\cite{Lincoln-Mitchell-Scedrov-Shankar-1992} % NP-completeness with weakening
\cite{Kanovich-1992} % NP-completeness for MLL
\cite{Lincoln-Winkler-1994} % NP-completeness for MLL without atoms

@@
